```{r setup7, include=FALSE} 
library(dada2)
library(tidyverse)
library(phyloseq)

```

# Phyloseq

The phyloseq R package is a powerful framework for further analysis of microbiome data. We now demonstrate how to straightforwardly import the tables produced by the `dada2` pipeline into phyloseq. 

## Import

```{r}
path <- "~/Desktop/microbiomeworkshop2022/"


## Read in files
seqtab <- readRDS(file.path(path, "results/seqtab_final.rds"))
taxa <- readRDS(file.path(path, "results/taxa_final.rds"))
```

### Exercise

1. Read the metadata file into R, called `info`. 

2. Phyloseq needs the rownames of the metadata to match the sample names. How can you accomplish this? 

```{r, echo = FALSE}
info <- read.table(file.path(path, "project_metadata.txt"), header = TRUE)

## Name rows after Sample ID
rownames(info) <- info$SampleID
```

Make a phyloseq object:

```{r}
ps <- phyloseq(otu_table(seqtab, taxa_are_rows=FALSE), sample_data(info), tax_table(taxa))
ps
```

In order to look at the ASVs, we need to remove the sequences.

```{r}
## Remove sequence names, rename to something manageable
asv_names <- vector(dim(otu_table(ps))[2], mode = "character")
for (i in 1:dim(otu_table(ps))[2]){
	asv_names[i] <- paste("ASV", i, sep = "_")
}
taxa_names(ps) <- asv_names
colnames(otu_table(ps)) <- asv_names
rownames(tax_table(ps)) <- asv_names
```

You can see that the phyloseq object has an `otu_table`(ASV table), `sample_data` and `tax_table`. You can use functions `tax_table()`, `sample_data()` and `otu_table()` to access the data.

### Exercise

Take a look the following functions and find out what they do:
- subset_samples()
- subset_taxa()
- tax_glom()
- sample_sums()
- prune_samples()
- transform_sample_counts()
- psmelt()

## Preprocess
